---
title: "Your Informative Title Here"
subtitle: "Case Study 2" 
author:
  - name: "Radiah Khan"
  - name: "Catherine Weeks" 
  - name: "Mercer Mercer"
  - name : "Nafisa Mohammad"
date: November 14
format: 
  pdf:
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
execute: 
  echo: true
  message: false
  warning: false
---

## Introduction

## Data Description

In this case study, we are working with the ex1029 data set available in the Sleuth2 library. This data set represents a sample of 25,632 male respondents to the March 1988 U.S. Current Population Survey that is administered monthly by the U.S. Census Bureau.

```{r, include= FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(flextable)
library(Sleuth2)
wages <- Sleuth2::ex1029
wages
```

### Data Wrangling

Before analysis, we checked how many missing observations were present in this data set. Using the `is.na()` and `colSums()` functions, we confirmed that there were no missing observations, so no rows were removed and therefore the dataset used in our analysis contains 25,632 complete observations.

```{r, include=FALSE}
#Check for n/a values
sum(is.na(wages))
colSums(is.na(wages))
```

To further prepare for visualization and modeling, a new categorical variable, `exp_group` was created by grouping `Experience` into four intervals: 0-20 years, 21-40 years, 41-60 years, and 60+ years of experience.

```{r, include=FALSE}
#Just has the experience group
wages_exp_group <- wages |>
  mutate(exp_group = case_when(
    Experience < 21 ~ "0–20",
    Experience < 41 ~ "20–40",
    Experience < 61 ~ "41-60",
  
    TRUE     ~ "60+"
  ))
wages_exp_group
```

To visualize regional differences, we computed the average weekly wage by region and race and the average education level by region and race.

```{r, include=FALSE}
# Wrangling the data
#Average wage in each regions by race
wages_region_avg_wage <- wages |>
  select(Wage, Black, Region) |>
  group_by(Region, Black) |>
  summarise(avg_wage_region = mean(Wage, na.rm = TRUE))|>
  ungroup()
wages_region_avg_wage
```

```{r, include=FALSE}
#EDA Potential 
# - (Wage and Race)
# - (Wage by regions)
# Have a combined plot on that if needed for the research question
#Need input on the EDA
```

```{r, include=FALSE}
#Exp group + region
wages_exp_group_region <- wages_exp_group |>
  mutate(Race = ifelse(Black == "Yes", "Black", "Non-Black")) |>
  group_by(Region, Race, exp_group) |>
  summarise(avg_wage = mean(Wage, na.rm = TRUE), .groups = "drop")
```

The visualization below shows how average wages increase with experience across all regions and that Black respondents consistently earn lower wages than non-Black respondents.

```{r, echo= FALSE}
#Plotting EDA (Exp group and avg wage by race)
#The region plot looks really condensed 
ggplot(wages_exp_group_region, 
       aes(x = exp_group, y = avg_wage, fill = Race)) +
  geom_col(position = position_dodge(width = 0.7)) +
  facet_wrap(~ Region) +
  labs(title = "Average Wage by Experience Group among Races",
       x = "Experience Group (years)",
       y = "Average Wage") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

#### Preparing for Regression Analysis

Before modeling, we converted our categorical variables to factors and set the "No" level of Black as the reference category.

```{r}
#factoring the categorical variables in the wage data set
wages$Black <- factor(wages$Black)
wages$Region <- factor(wages$Region)

#setting the new reference level
wages$Black <- relevel(wages$Black, ref = "No")
```

```{r, include=FALSE}
#MLM model
mlm_model_untransformed <- lm(Wage ~ Education  + Experience + Black * Region, data = wages)
summary(mlm_model_untransformed)
```

```{r, echo=FALSE}
#Prettifying it by adding tables : Making it a data frame first to table it
mlm_untransformed_df <- tidy(mlm_model_untransformed, conf.int = TRUE)
flextable(mlm_untransformed_df)|>
  set_caption("Regression Summary of the Untransformed Model: Predicting Wage")|>
  autofit()|>
  theme_apa()
```

```{r}
# Checking Linearity and Equal Variance
ggplot(mlm_model_untransformed |> augment(), aes(x = .fitted, y = .resid))+
  geom_point()+
  geom_hline(yintercept = 0, col = "red")+
  labs(title = "Residual Plot (Untransformed Model)" ,
       x = "Fitted Value",
       y = "Residuals")+
  theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

```{r}
# Checking for the normality condition
mlm_model_untransformed |>
  augment() |> ggplot(aes(sample = .resid)) + geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot (Untransformed Model) ", x = "Quantilies", y = "Sample Outcomes") + 
  theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r, echo= FALSE}
#MLM model with log scaled y axis
mlm_model <- lm(log(Wage) ~ Education  + Experience + Black * Region, data = wages)
summary(mlm_model)

```

```{r}
#Prettifying it 
mlm_transformed_df <- tidy(mlm_model, conf.int = TRUE)
flextable(mlm_transformed_df)|>
  set_caption("Regression Summary of the Transformed Model: Predicting Wage")|>
  autofit()|>
  theme_apa()
```

```{r}
# Checking Linearity and Equal Variance
ggplot(mlm_model |> augment(), aes(x = .fitted, y = .resid))+
  geom_point()+
  geom_hline(yintercept = 0, col = "red")+
  labs(title = "Residual Plot (Transformed Model)",
       x = "Fitted",
       y = "Residuals") + 
  theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r}
# Checking for the normality condition
mlm_model |>
  augment() |> ggplot(aes(sample = .resid)) + geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot (Transformed model) ", x = "Quantilies", y = "Sample Outcomes") + 
  theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r, include= FALSE}
wages_reduced_model <- lm(log(Wage) ~ Education + Experience + Black + Region, data = wages)
summary(wages_reduced_model)

```

```{r, echo=FALSE}
wages_reduced_df <- tidy(wages_reduced_model, conf.int = TRUE)
flextable(wages_reduced_df)|>
  set_caption("Regression Summary of the Reduced Model: Predicting Wage")|>
  autofit()|>
  theme_apa()
```

```{r}
f_results <- anova(wages_reduced_model,mlm_model)
f_results
```

```{r}
anova_df <- as.data.frame(f_results)
flextable(anova_df)|>
  set_caption("F-test Result Summary")|>
  autofit()|>
  theme_apa()
```

```{r}
#creating new labels for the facet wrapped variables 
graph.labels <- c("Yes"="Black","No" = "Non-Black")

#plotting a graph exploring Avg pay by region faceted by race
wages_region_avg_wage |> ggplot(aes(x=avg_wage_region, y = Region, fill = Black)) + 
  geom_col()+ 
  facet_grid(~Black,labeller = labeller(Black = graph.labels) )+
  coord_flip() +labs(title = "Average Pay in Each USA Region by Race", x= "Average Weekly Wage")+
  theme_bw()
```

#Region does not affect wages in races --\> then the question is not if race affects wage, it is if region affects the wage gap

```{r}
library(knitr)
#Saving summary table of wages data
wages_summary <-summary(wages)
wages_df <- as.data.frame(wages_summary)


#printing the summary table created above with captions.
#flextable(wages_df)|>
  #set_caption("Wage Summary")|>
  #autofit()|>
  #theme_apa()
```

Graph for Average Education by Region and Race, USA

```{r}

#get averages of education per region
wages_edu_group_region <- wages |>
  group_by(Region, Black) |>
  summarise(avg_education = mean(Education)) |>
  ungroup()

graph.labels <- c("Yes"="Black","No" = "Non-Black")

ggplot(wages_edu_group_region, aes(x = avg_education, y = Region, fill = Black)) +
  geom_col() +
  facet_grid(~Black,labeller = labeller(Black = graph.labels) )+
  labs(
    title = "Average Education by Region and Race, USA",
    x = "Average Education (in Years of Schooling)",
    y = "Region"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.line.y = element_blank(),
    axis.line.x = element_blank(),
  ) + coord_flip()
```
